Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd
Went over the spellcheck to fix many typos

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"

filters:
  - acronyms
  - highlight-text
  - abstract-section

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          % NOTE: make siunitx's \qty work without clashing with the one in physics
          \let\qty\SI
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html

author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium

bibliography: media/references.bib

acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml

---

# Abstract
<!-- teaser --> The recent decades observed an increasing number of space activities resulting in higher numbers of re-entries at the end of the mission lifespan or orbital decay. 
The risk of collision between uncontrolled re-entering space debris and aircraft
is  increasing due to both ever-growing number of airborne flights and 
of orbiting bodies (and relevant re-entries). 
<!-- why important --> National and international authorities need to balance safety and economic
impact of adopted measures should a re-entry event be announced, usually on short
notice. <!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial flight densities calculated from
trajectories as recorded by EUROCONTROL's Network Manager and
binned with the Uber H3 geospatial indexing system.  
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the model by employing using actual flown flight trajectories as a basis for time-varying flight densities rather than relying on proxies such as the population density or ADS-B/Mode S message counts.
These densities combined with a buffered area along the eventual path of
re-entry, can provide mid-air impact probabilities to authorities as an aid
for deciding on safety measures such as airspace closures and/or re-routings.

@@

<!-- teaser opening --><!-- might need a rewrite - first jotting down -->
The quest for space is a driver for technological development.
Key milestones include the first human traval to space and surrounding the globe, the first and now many succeeding successful landings on the moon, and regular operations of space stations in the orbit.
These developments helped to identify options for the wider use of space, benefits, and its commercialisation.
This is showcased by an increasing number of activities aimed to generate value through the exploration of space ranging from scientific understanding and experimentation, sustainable prooduction of goods in a non-gravity environment, services, and recently leisure.
The increased number of launches, however, also increased the number of artefacts <!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->
With forecasts predicting a steadily growing number of flights, see @fig-forecast, and the modern
race to commercialise space, there is a need to re-evaluate the risks of
collision and casualties due to uncontrolled debris re-entries.
The increasing risk landscape poses challenges in terms of oversight and risk management.
There is a growing demand by overseeing authorities for new methodologies and tools to assess which 
safety measures are the most adequate to minimise potential casualties and economic
impact.

The increase in space operations does not exist in a vacuum. 
The continual growth of air traffic - despite temporal bumps - is evidenced by the surging number of flights in Europe [@forecastUpdate2025] and across the world. 
On-going development/deployment activities of mega-constellations of satellites in Low Earth Orbit \acr{LEO}, 
such as already active Starlink and OneWeb or planned ones (e.g. Kuiper, Starshield, Xingwang) will add thousand of orbiting objects with a short operational life. 
Some of these will inevitably re-enter the Earth's atmosphere in an uncontrolled manner, c.f. @tbl-megaconstellations [@megaconstellations].

@@

<!-- contribution of the paper -->
This paper develops a practical and straightforward data-driven pipeline to rapidly
assess the risk from uncontrolled re-entry of space debris for the European
region.
Apart from the general risk assessment of potential re-entries, 
a soon to occur re-entry of debris from a 1972 failed USSR mission to Venus is analysed.

::: {#fig-forecast}

```{r forecast}
#| out-width: "100%"
#| echo: false
#| fig-scap: "Forecast"
knitr::include_graphics("media/figures/forecast_statfor.png")
```

@@

```{r}
#| label: tbl-megaconstellations
#| tbl-cap: Megaconstellations (planned > 1000) as per filing to the \acr{ITU}
#| tbl-cap-location: bottom

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->
The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or rotation of re-entering objects increase the level of uncertainty for modelling.
Common approaches employ a mix of handling the uncertainty of object or debris movement, the associated exposed area, including temporal uncertainty windows.

      <!-- is there more to say and provide context? -->

<!-- APPROACH / METHOD --------------------------------------------------- -->
<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->
<!-- as a minimum, we should say something about the WHAT -->
In general, the method for estimating the risk to commercial aircraft from reentering debris objects revolves around estimating the exposure of a single flight and the population of flights to the re-entering object.
In that respect, a robust determination of flight density is a key aspect.

@@

::: {#fig-cells}

```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

H3 cells (red) at resolution 3 and their centers (blue) covering the bounding
box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded polygon).

:::

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>
This paper expands the works of [@wrightAirspaceClosuresDue2025; @patera2008] and provides a reproducible example for States, Agencies or relevant authorities to build upon.

<!-- next sentence feels like an orphan ... do we need this here? ------>
Next sections will develop the mathematical background for the evaluation of collision and casualties risk of which traffic density is a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of July 2024.
The chart shows the expected outcomes in terms of major traffic flows, and areas of high or low traffic.
For example, cells with a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libia or Syria). The major traffic flows (North Atlantic, Canaries
and East-bound over Türkiye) are clearly highlighted showing densities of more than 10.

::: {#fig-density-summer2024-res-3}

```{r density-winter-res-3}
#| out-width: "110%"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

where 
* $\rho({\vec{x}})$ is the probability density of an aircraft in $\vec{x}$,
* $\sigma(\vec{x})$ is the impact probability density of a debris object $k$
in the position $\vec{x}$, and 
* $\varepsilon_k$ is the effective exposed area, i.e., the effective risk area for the $k^{th}$ debris object impacting
the aircraft (see @sec-eea later on).

Our approach builds on the discretisation of the geo-spatial risk area.
For this purpose we use a hexagon-based tesselation pattern.
Uber's H3 method ensures non-overlapping hexagons, with an equidistant relationship between all neighbours, and enables an efficient indexing of the tessel-cells. 
Using H3 to discretise the area of interest $a$ gives

@@

## Aircraft effective exposed area{#sec-eea}

For the actual collision encounter between a re-entering object and an aircraft, we assume the aircraft impact area to be much larger than the debris object area.
Thus, the area spanned by the debris object is smaller than the aircraft dimensions. 
This assumption holds for practically all commercial aircraft within the scope of this study, and we can neglect the debris impact area.
Accordingly, we can also simplify the geometry of the aircraft and model it as a rectangular solid with dimesions of wingspan ($W$), length ($L$) and height ($H$).

Regarding the aircraft-debris encounter geometry, this yields to two conceptual considerations. 
<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front plane? Does the literture use "view"  -->
1.) if the aircraft were stationary, the area exposed to the debris impact would be the top view, and
2.) if the debris were stationary, it would be the front view.

Our focus is on commercial aircraft in flight, and - implicitly - operating at higher altitudes.
The en-route portion of a flight is characterised by the aircraft in straight and level flight, moving (mainly) horizontally.
At the same time, the movement of the re-entered debris object at regular en-route flight levels will be vertically [^balistic_descent].
This encounter geometry allows to express the effective exposed area, $\varepsilon_k$, as

@@

where 
* $s_{a}$ is the average aircraft speed, 
* $s_{d_k}$ is the average fall speed of the $k$-th debris object, 
* $a_T$ area of the aircraft as viewed from the top, and 
* $a_F$ area of the aircraft as viewed from the from the front of the aircraft.

@@

* constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$
* constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed, 
* aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$, 
  where $W_t$ is the wingspan, and $L_t$ the length for the specific aircraft
  type $t$
* aircraft area as viewed from the front, $a_F = W_t \times H_t = F_t$, 
  where $W_t$ is the wingspan, and $H_t$ the height for the specific aircraft
  type $t$

@@

These aircraft properties are well-known. 
While there does not exist a single openly available date base for these aircraft parameters, we collected the information from a series of sources and integrated them into a lookup dataset.
The values $W_t$, $H_t$ and $L_t$ have been retrieved from [@performance_database;@skybrary;@doc8643].

@@

[^balistic_descent]: *Even for a ballistic coefficient of \qty{50}{lbs/ft^2},
the flight path angle of a reentering object at \qty{40000}{ft} is 
\ang{-86.6}, which corresponds to nearly vertical descent.
Since most debris objects have ballistic coefficients equal to or 
less than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
accurate approximation.* [@patera2008]

@@

Debris that reenter the Earth’s atmosphere via drag-induced decay will fly
circular orbits prior to re-entry even if initially on elliptical
ones.[^circular-orbit]

The angular position, $\theta$, of the re-entry point from the ascending node, 
see @sec-orbitals, is uniformly distributed for uncontrolled rentries,
$\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in atmospheric drag
and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformely distributed.

The impact probability density, $\sigma(\varphi, i)$,
at latitude $\varphi$ of a reentring object whose orbit has inclination $i$
(see @fig-orbital-elements) is given by [[Equation 15 in @patera2008]]:

@@

where $R$ is the distance from the center of the Earth of the object potentially
being impacted by the debris.

So for example if a building is located at latitude 30 degrees and has an area 
of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth’s atmosphere
from an orbit having an inclination of 45 degrees, the impact probability is 
given by

@@

where we used the equatorial value of $\qty{6378}{km}$, for the Earth's radius,
$R$, as the distance of the building from the center of the planet.

@@

::: {#fig-weighting-inclination-i}

```{r weighting-inclination-i}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

@@

[^circular-orbit]: In fact drag at perigee will lower apogee and similarly drag
at apogee will lower perigee. But because drag at perigee is higher due to
denser atmosphere, the apogee  will decrease faster than perigee and the orbit
will then tend to be a circular one.

@@

::: {#fig-weighting}

```{r weighting}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting.png")
```

@@

::: {#fig-weighting-nm-area}

```{r weighting-nm-area}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_on_Earth.png")
```

@@

where 
* $E_{h,H}$ is the collision expectation in the hexagon $h$ at hour $H$,
* $W(h)$ is the median value of the weighting function over the area of hexagon
$h$,
* $\rho_{t,H}(h)$ is the traffic density for the \acr{ICAO} aircraft type $t$ in
cell $h$ at hour $H$, and
* $\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

::: {#fig-collision}

```{r collision}
#| out-width: "100%"
#| echo: false
knitr::include_graphics(here::here("media", "figures", "collision_expectation_2024-07-05_3.png"))
```

@@

# Application Use Case: KOSMOS 482 Descent Craft re-entry
<!-- I would make this a separate section in the paper -->

@@

::: {#fig-kosmos}

```{r kosmos}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/example-image.png")
```

@@

We have calculated collision and casuality risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities is available in the companion
GitHub repository of the paper at <https://github.com/espinielli/flight_density_space_debris>.
We have also developed a companion R package `aviodebris` to heavylift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/espinielli/aviodebris>.

::: {#fig-data-densities}

```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

<!-- opening statement linking back to abstract/intro and movtivation for this work --------->
Space objects and debris re-entry is an increasing risk to aviation and the general public.
The higher pace of commercial space operations will see an increase of launches and subsequent re-entries.
Dependent on the mission, objects re-entering the atmosphere can take on differnt forms, shapes, and sizes.
On top, a higher number of these entries may be uncontrolled.
With the continual growth of air traffic, there is a need to enhance aviation safety risk awareness and respond in a more tailored fashion to accomodate the associated uncertainties, reduce the impact on aviation, and limit the risk to the travelling public.

<!-- what was shown in the paper ------------------------------------------------------------>
This paper expands on previous work.
With a focus on Europe, we establish an upper bound for the traffic density based on actual flights.

@@

The approach presented in this paper builds on previous research.
Our approach utilises H3-tesselation to reduce the spatial complexity.
This cell indexing approach reduces the computation significantly.

        WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.

@@

<!-- application of the results -------------------------------------------------------------->
We show how National and/or International authorities could prepare for the more and more likely and frequent occurrence of uncontrolled re-entry of space objects that could pose a safety concern on civilian aviation.
The techniques we implemented on realistic data and sound algorithms can contribute to the bag of tools in the hands of technical and political entities aiming at guaranteeing the utmost safety and minimal economic disruption to civil aviation in Europe.

As part of this research and to inform the decision-makers it is planned to explore the impact of megacostellations and flight growth forecast and what would happen in the mid- to long-term in varios scenarios of uncontrolled re-entries and the risk for European airspace.

@@

:::{#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}

```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semiparameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

Definition of Keplerian elements. The orbital elements describing the trajectory of orbiting object are illustrated in association with an inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis, $\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$, is perpendicular to the $I$-axis in the equatorial plane n the sense of Earth's rotation with the $K$-axis pointing up. The semi-major axis, $a$, the eccentricity, $e$, the inclination, $i$, the right ascension of the ascending node, $\Omega$ , and the argument of perigee, $\omega$, constitute, together with the mean anomaly at the initial position, $M_0$, a set of six constants of motion that are preserved in the absence of perturbation. True anomaly, $\nu$, specifies the location of an orbiting object within the orbit. It is the angle between perigee and the orbiting object’s position vector measured in the direction of the orbiting object’s motion.
Figures from [@sellers2003understanding], respectively Figure 5-2 and Figure 5-9.

:::

@@

These are used to derive numerically compute the weighting function,
@fig-weighting-inclination-i:
for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
positions and calculate the corresponding cartesian position and the
equivalent longitude and latitude.

>>>>

---
title: "Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects"
filters:
  - acronyms
  - highlight-text
  - abstract-section
format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          \let\qty\SI
          % NOTE: make siunitx's \qty work without clashing with the one in physics
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html
author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium
bibliography: media/references.bib
acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml
editor: 
  markdown: 
    wrap: 72
---
```{r}
```

```{r}
```

# Abstract

<!-- teaser -->

The recent decades observed an increasing number of space activities
resulting in higher numbers of re-entries at the end of the mission
lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to both
ever-growing number of airborne flights and of orbiting bodies (and
relevant re-entries). <!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures should a re-entry event be announced, usually on short notice.
<!-- no hard line breaks in abstracts -->
<!-- approach / experiments --> We employ hourly and daily spatial
flight densities calculated from trajectories as recorded by
EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
indexing system.\
Impact risk probabilities are then calculated based on previous
theoretical work. <!-- results / conclusions --> This work enhances the
model by using actual flown flight trajectories as a basis for
time-varying flight densities rather than relying on proxies such as the
population density or ADS-B/Mode S message counts. These densities
combined with a buffered area along the eventual path of re-entry, can
provide mid-air impact probabilities to authorities as an aid for
deciding on safety measures such as airspace closures and/or
re-routings.

@@

<!-- teaser opening -->

<!-- might need a rewrite - first jotting down -->

The quest for space is a driver for technological development. Key
milestones include the first human travel to space and surrounding the
globe, the first and now many succeeding successful landings on the
moon, and regular operations of space stations in the orbit. These
developments helped to identify options for the wider use of space,
benefits, and its commercialisation. This is showcased by an increasing
number of activities aimed to generate value through the exploration of
space ranging from scientific understanding and experimentation,
sustainable production of goods in a non-gravity environment, services,
and, more recently, space tourism. The increased number of launches,
however, also increased the number of artefacts
<!-- falling down and --> re-entering the atmosphere.

<!-- zooming in on problem -->

With forecasts predicting a steadily growing number of flights, see
@fig-forecast, and the modern race to commercialise space, there is a
need to re-evaluate the risks of collision and casualties due to
uncontrolled debris re-entries. The increasing risk landscape poses
challenges in terms of oversight and risk management. There is a growing
demand by overseeing authorities for new methodologies and tools to
assess which safety measures are the most adequate to minimise potential
casualties and economic impact.

The increase in space operations does not exist in a vacuum. The
continual growth of air traffic - despite temporal bumps - is evidenced
by the surging number of flights in Europe [@forecastUpdate2025] and
across the world. On-going development/deployment activities of
mega-constellations of satellites in Low Earth Orbit \acr{LEO}, such as
already active Starlink and OneWeb or planned ones (e.g. Kuiper,
Starshield, Xingwang) will add thousand of orbiting objects with a short
operational life. Some of these will inevitably re-enter the Earth's
atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
@megaconstellations].

@@

<!-- contribution of the paper -->

This paper develops a practical and straightforward data-driven pipeline
to rapidly assess the risk from uncontrolled re-entry of space debris
for the European region. Apart from the general risk assessment of
potential re-entries, a soon to occur re-entry of debris from a 1972
failed USSR mission to Venus is analysed.

::: {#fig-forecast}
```{r forecast}
#| out-width: "100%"
#| echo: false
#| fig-scap: "Forecast"
knitr::include_graphics("media/figures/forecast_statfor.png")
```

@@

```{r}
#| label: tbl-megaconstellations
#| tbl-cap: Mega-constellations (planned > 1000) as per filing to the \acr{ITU}
#| tbl-cap-location: bottom

@@

<!-- ESTABLISH SOME CONTEXT ---------------------------------------------- -->

The modelling and tracking of uncontrolled re-entries poses challenges.
Actual and historic data is sparse, and the size, shape, and attitude or
rotation of re-entering objects increase the level of uncertainty for
modelling. Common approaches employ a mix of handling the uncertainty of
object or debris movement, the associated exposed area, including
temporal uncertainty windows.

```         
  <!-- is there more to say and provide context? -->
```

<!-- APPROACH / METHOD --------------------------------------------------- -->

<!-- it would be cool, if we can add a diagram here that shows the path from data, major processing building blocks to the results -->

<!-- as a minimum, we should say something about the WHAT -->

In general, the method for estimating the risk to commercial aircraft
from reentering debris objects revolves around estimating the exposure
of a single flight and the population of flights to the re-entering
object. In that respect, a robust determination of flight density is a
key aspect.

@@

::: {#fig-cells}
```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

H3 cells (red) at resolution 3 and their centres (blue) covering the
bounding box of 43 EUROCONTROL Member States \acr{FIR}s (blue shaded
polygon).
:::

@@

<!-- next sentence feels like an orphan ... do we need this here? ------>

This paper expands the works of [@wrightAirspaceClosuresDue2025;
@patera2008] and provides a reproducible example for States, Agencies or
relevant authorities to build upon.

<!-- next sentence feels like an orphan ... do we need this here? ------>

Next sections will develop the mathematical background for the
evaluation of collision and casualties risk of which traffic density is
a key component.

@fig-density-summer2024-res-3 shows hourly flight density for the 5th of
July 2024. The chart shows the expected outcomes in terms of major
traffic flows, and areas of high or low traffic. For example, cells with
a density of zero are clearly visible for areas of war or no-fly zones
(Ukraine, Libya or Syria). The major traffic flows (North Atlantic,
Canaries and East-bound over Türkiye) are clearly highlighted showing
densities of more than 10.

::: {#fig-density-summer2024-res-3}
```{r density-winter-res-3}
#| out-width: "110%"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

where \* $\rho({\vec{x}})$ is the probability density of an aircraft in
$\vec{x}$, \* $\sigma(\vec{x})$ is the impact probability density of a
debris object $k$ in the position $\vec{x}$, and \* $\varepsilon_k$ is
the effective exposed area, i.e., the effective risk area for the
$k^{th}$ debris object impacting the aircraft (see @sec-eea later on).

Our approach builds on the discretisation of the geospatial risk area.
For this purpose we use a hexagon-based tessellation pattern. Uber's H3
method ensures non-overlapping hexagons, with an equidistant
relationship between all neighbours, and enables an efficient indexing
of the tessel-cells. Using H3 to discretise the area of interest $a$
gives

@@

## Aircraft effective exposed area {#sec-eea}

For the actual collision encounter between a re-entering object and an
aircraft, we assume the aircraft impact area to be much larger than the
debris object area. Thus, the area spanned by the debris object is
smaller than the aircraft dimensions. This assumption holds for
practically all commercial aircraft within the scope of this study, and
we can neglect the debris impact area. Accordingly, we can also simplify
the geometry of the aircraft and model it as a rectangular solid with
dimensions of wingspan ($W$), length ($L$) and height ($H$).

Regarding the aircraft-debris encounter geometry, this yields to two
conceptual considerations.
<!-- TERMINOLOGY? rectangular solid -- impact --> top plane, front
plane? Does the literature use "view" --\> 1.) if the aircraft were
stationary, the area exposed to the debris impact would be the top view,
and 2.) if the debris were stationary, it would be the front view.

Our focus is on commercial aircraft in flight, and - implicitly -
operating at higher altitudes. The en-route portion of a flight is
characterised by the aircraft in straight and level flight, moving
(mainly) horizontally. At the same time, the movement of the re-entered
debris object at regular en-route flight levels will be vertically [^2].
This encounter geometry allows to express the effective exposed area,
$\varepsilon_k$, as

[^2]: *Even for a ballistic coefficient of* \qty{50}{lbs/ft^2}, the
    flight path angle of a reentering object at \qty{40000}{ft} is
    \ang{-86.6}, which corresponds to nearly vertical descent. Since
    most debris objects have ballistic coefficients equal to or less
    than \qty{50}{lbs/ft^2}, the vertical descent assumption is a very
    accurate approximation. [@patera2008]

@@

where \* $s_{a}$ is the average aircraft speed, \* $s_{d_k}$ is the
average fall speed of the $k$-th debris object, \* $a_T$ area of the
aircraft as viewed from the top, and \* $a_F$ area of the aircraft as
viewed from the from the front of the aircraft.

@@

-   constant debris fall speed $s_{d_k} = \qty{145}{mph} = S$ [[page 8
    @patera2008]\]
-   constant aircraft speed, $s_{a} = C_t$, equal to its cruise speed,
-   aircraft area as viewed from the top, $a_T = W_t \times L_t = A_t$,
    where $W_t$ is the wingspan, and $L_t$ the length for the specific
    aircraft type $t$
-   aircraft area as viewed from the front,
    $a_F = W_t \times H_t = F_t$, where $W_t$ is the wingspan, and $H_t$
    the height for the specific aircraft type $t$

@@

These aircraft properties are well-known. While there does not exist a
single openly available date base for these aircraft parameters, we
collected the information from a series of sources and integrated them
into a lookup dataset. The values $W_t$, $H_t$ and $L_t$ have been
retrieved from [@performance_database; @skybrary; @doc8643]; we used
\qty{1000}{m^2} for $\varepsilon_t$ in case of missing aircraft type.

@@

@@

Debris that reenter the Earth's atmosphere via drag-induced decay will
fly circular orbits prior to re-entry even if initially on elliptical
ones.[^3]

[^3]: In fact drag at perigee will lower apogee and similarly drag at
    apogee will lower perigee. But because drag at perigee is higher due
    to denser atmosphere, the apogee will decrease faster than perigee
    and the orbit will then tend to be a circular one.

The angular position, $\theta$, of the re-entry point from the ascending
node, see @sec-orbitals, is uniformly distributed for uncontrolled
reentries, $\sigma(\theta) = 1 / (2\pi)$. This is due to uncertainties in
atmospheric drag and the gravitational effects of the Equatorial bulge.
Hence we assume that the impact longitude, $\lambda$, is uniformly
distributed.

The impact probability density, $\sigma(\varphi, i)$, at latitude
$\varphi$ of a reentering object whose orbit has inclination $i$ (see
@fig-orbital-elements) is given by [[Equation 15 in @patera2008]\]:

@@

where $R$ is the distance from the centre of the Earth of the object
potentially being impacted by the debris.

So for example if a building is located at latitude 30 degrees and has
an area of $\qty{1000}{m^2}$ and a piece of debris reenters the Earth's
atmosphere from an orbit having an inclination of 45 degrees, the impact
probability is given by

@@

where we used the equatorial value of $\qty{6378}{km}$, for the Earth's
radius, $R$, as the distance of the building from the centre of the
planet.

@@

::: {#fig-weighting-inclination-i}
```{r weighting-inclination-i}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

@@

@@

::: {#fig-weighting}
```{r weighting}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting.png")
```

@@

::: {#fig-weighting-nm-area}
```{r weighting-nm-area}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_on_Earth.png")
```

@@

where \* $E_{h,H}$ is the collision expectation in the hexagon $h$ at
hour $H$, \* $W(h)$ is the median value of the weighting function over
the area of hexagon $h$, \* $\rho_{t,H}(h)$ is the traffic density for
the \acr{ICAO} aircraft type $t$ in cell $h$ at hour $H$, and \*
$\varepsilon_t$ is effective exposed area for aircraft of type $t$.

@@

::: {#fig-collision}
```{r collision}
#| out-width: "100%"
#| echo: false
knitr::include_graphics(here::here("media", "figures", "collision_expectation_2024-07-05_3.png"))
```

@@

# Application Use Case: KOSMOS 482 Descent Craft re-entry

<!-- I would make this a separate section in the paper -->

@@

::: {#fig-kosmos}
```{r kosmos}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/example-image.png")
```

@@

We have calculated collision and casuality risk probabilities for the
two busiest weeks in 2023 and 2024. The relevant dataset represents a
data cube as logically depicted in @fig-data-densities is available in
the companion GitHub repository of the paper at
<https://github.com/espinielli/flight_density_space_debris>. We have
also developed a companion R package `aviodebris` to heavy lift the
various data processing steps to produce the results in the paper; it is
available at <https://github.com/espinielli/aviodebris>.

::: {#fig-data-densities}
```{r densities}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/cubes.png")
```

@@

<!-- opening statement linking back to abstract/intro and motivation for this work --------->

Space objects and debris re-entry is an increasing risk to aviation and
the general public. The higher pace of commercial space operations will
see an increase of launches and subsequent re-entries. Dependent on the
mission, objects re-entering the atmosphere can take on different forms,
shapes, and sizes. On top, a higher number of these entries may be
uncontrolled. With the continual growth of air traffic, there is a need
to enhance aviation safety risk awareness and respond in a more tailored
fashion to accommodate the associated uncertainties, reduce the impact
on aviation, and limit the risk to the travelling public.

<!-- what was shown in the paper ------------------------------------------------------------>

This paper expands on previous work. With a focus on Europe, we
establish an upper bound for the traffic density based on actual
flights.

@@

The approach presented in this paper builds on previous research. Our
approach utilises H3-tessellation to reduce the spatial complexity. This
cell indexing approach reduces the computation significantly.

```         
    WHAT ARE THINGS THAT ARE "WEAKNESSES" = STUFF THAT CAN BE IMPROVED? IT WOULD BE GOOD TO HAVE A LINE OR TWO ON THIS.
```

@@

<!-- application of the results -------------------------------------------------------------->

We show how National and/or International authorities could prepare for
the more and more likely and frequent occurrence of uncontrolled
re-entry of space objects that could pose a safety concern on civilian
aviation. The techniques we implemented on realistic data and sound
algorithms can contribute to the bag of tools in the hands of technical
and political entities aiming at guaranteeing the utmost safety and
minimal economic disruption to civil aviation in Europe.

As part of this research and to inform the decision-makers it is planned
to explore the impact of mega-constellations and flight growth forecast
and what would happen in the mid- to long-term in various scenarios of
uncontrolled re-entries and the risk for European airspace.

@@

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

Definition of Keplerian elements. The orbital elements describing the
trajectory of orbiting object are illustrated in association with an
inertial reference frame $\hat{I}$-$\hat{J}$-$\hat{K}$. The $I$-axis,
$\hat{I}$, points toward the vernal equinox and the $J$-axis, $\hat{J}$,
is perpendicular to the $I$-axis in the equatorial plane n the sense of
Earth's rotation with the $K$-axis pointing up. The semi-major axis,
$a$, the eccentricity, $e$, the inclination, $i$, the right ascension of
the ascending node, $\Omega$ , and the argument of perigee, $\omega$,
constitute, together with the mean anomaly at the initial position,
$M_0$, a set of six constants of motion that are preserved in the
absence of perturbation. True anomaly, $\nu$, specifies the location of
an orbiting object within the orbit. It is the angle between perigee and
the orbiting object's position vector measured in the direction of the
orbiting object's motion. Figures from [@sellers2003understanding],
respectively Figure 5-2 and Figure 5-9.
:::

@@

These are used to derive numerically compute the weighting function,
@fig-weighting-inclination-i: for debris with inclination $i$ we
simulate an orbit of $N$, i.e. $10001$, positions and calculate the
corresponding Cartesian position and the equivalent longitude and
latitude.
